Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-05-07 WSSSPE2_report/WSSSPE2_review.tex
Section 4: clean up, small edits, and format changes.

\subsection{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable
Software through Sustainable Software Architectures~\cite{Venters_poster}}
Venters proposed that sustainable software is a composite, first-class,
non-functional requirement (NFR) that is at a minimum a measure of a system's
maintainability and extensibility, but may also include other NFRs such as
efficiency (e.g., energy, cost), interoperability, portability, reusability,
scalability, and usability. To achieve technically sustainable software, Venters
suggested that software architectures are fundamental as they are the primary
carrier of system NFRs, i.e., pre-system understanding; and influence how
developers are able to understand, analyze, extend, test, and maintain a
software system, i.e., post-deployment system understanding. In addition,
Venters highlighted that sustainability of software architectures needs to be
addressed to endure different types of change and evolution in order to mitigate
architectural drift, erosion, and knowledge vaporization.

@@

\subsection{Marlon Pierce: Patching It Up, Pulling It
Forward~\cite{Pierce_poster}} Pierce discussed how open open source is. Open
software needs a diverse, openly governed community behind it, just as it needs
open licensing and a public code repository. To probe the level of governance
within open source projects, he and his co-authors (Suresh Marru and Chris
Mattmann) suggested a contest to encourage individual developers to submit
patches and pull request to projects that are important to them. This simple
mechanism will expose several governance mechanisms, such as how easy it is for
independent developers to communicate with project leadership, how projects
accept and properly license third-party contributions, and how projects make
decisions such as granting source tree write access.

\subsection{John Peterson: Continuous Integration for Concurrent MOOSE Framework
and Application Development on GitHub~\cite{Peterson_poster}} %Affiliation:
Idaho National Laboratory Peterson from the Idaho National Laboratory reported
that in March 2014, the MOOSE framework was released under an open source
license on GitHub, significantly expanding and diversifying the pool of current
active and potential future contributors on the project. The MOOSE team employs
an extensive continuous integration test suite to ensure that both the framework
and the applications based on the framework are verified before any code changes
are accepted into the repository. They use a combination of built-in Git
features such as branching and submodules, GitHub API integration capabilities,
and in-house developed testing software to perform this verification and update
the dependent applications in a relatively seamless manner for users.

\subsection{Abani Patra: Execute it~\cite{Patra_poster}}
%Univ at Buffalo, SUNY
Patra discussed the value of an available and easily accessible platform for
executing scientific software, e.g., HUBzero, to access XSEDE or other computing
resources. Such a platform for executing benchmark problems (even at small
scale) allows the developer community access to a reference implementation and
provides an easy way to train the larger user community. A second idea of this
presentation was that for true usability, much more attention and support needs
to be given to the integrated use of simulation tools inside complex workflows.
%needs supporting the user community workflow.

@@

\subsection{Daniel Katz: Implementing Transitive Credit with
JSON-LD~\cite{Katz_transitive_credit_poster}} Science and engineering research
increasingly relies on activities that facilitate research but are not rewarded
or recognized, such as: data sharing; developing common data resources, software
and methodologies; and annotating data and publications. To promote and advance
these activities, we must develop mechanisms for assigning credit, facilitate
the appropriate attribution of research outcomes, devise incentives for
activities that facilitate research, and allocate funds to maximize return on
investment. Katz discussed the issue of assigning credit for both direct and
indirect contributions by using JSON-LD to implement a prototype transitive
credit
system.

\subsection{Samin Ishtiaq: Daemons, Notifications and Sustaining Software}%~\cite{Ishtiaq_poster}}
%  Microsoft Research Cambridge
The reproduction and replication of novel results has become a major issue in
computer science, systems biology, and other computational disciplines. These
include both the inability to re-implement novel algorithms and approaches, and
lack of an agreement on how and what to benchmark these algorithms on. Ishtiaq
from Microsoft Research Cambridge pointed out these problems and made several
suggestions to address them.

\subsection{James Howison: Retract all Bit-Rotten Publications}%~\cite{Howison_poster}}
Howison sought to provoke discussion by proposing that papers whose workflows
are not kept current with the changing software ecosystem should be
automatically retracted. This would create an incentive for authors to keep
their software current and usable, rather than the current situation in which
every potential user has to do this individually. A softer version of the
proposal would identify papers whose software workflow has become bit-rotten and
allow others to keep the code up to date, either adding them as new authors of
the paper or providing credit in some other form for their academic service.

@@

\subsection{Robert Downs: Community Recommendations for Improving Sustainable
Scientific Software Practices~\cite{Downs_poster}} Robert Downs, of the Columbia
University Center for International Earth Science Information Network (CIESIN),
described a focus group study conducted with the Science Software Cluster (SSC)
of the Federation of Earth Science Information Partners (ESIP). For the study,
almost 300 attendees of the 2014 Summer ESIP Meeting were invited to participate
in simultaneous roundtable discussions on sustainability of science software.
Over two-thirds of the roundtable focus groups responded to a semi-structured
survey that contained three sets of questions eliciting recommendations for
near-term actions of the community to improve sustainable software practices.
Initial analysis of the participants' responses to the questionnaire revealed
several suggestions, which included improving community engagement and
collaborative activities, increasing understanding and awareness, and creating
incentives to motivate sustainable science software practices. The ESIP SSC
plans to engage the community in the recommended activities for improving
sustainable scientific software practices.

\begin{comment}
Downs RR, Lenhardt WC, Robinson E, Davis E, Weber N. 2014. Community
Recommendations for Improving Sustainable Scientific Software Practices. 2nd
Workshop on Sustainable Software for Science: Practice and Experiences
(WSSSPE2), New Orleans, Louisiana, 16 November 2014.
http://wssspe.researchcomputing.org.uk/wp-content/uploads/2014/11/Downs.pdf and
http://dx.doi.org/10.7916/D8Q52NBC
\end{comment}

\subsection{Carl Boettiger: rOpenSci: Building Sustainable Software by Fostering a
Diverse Community~\cite{Boettiger_poster}}
Boettiger described how the rOpenSci project has been successful by focusing not
just on building software but also on building a community of researchers who
learn and adopt their approaches to reproducible research and sustainable
software practice. Through outreach, mentoring, workshops, and hackathons, they
have not only reached new users, but also turned users into co-developers of
robust software and good practices to support data science research across a
growing set of disciplines.

\subsection{Jakob Blomer: The Need for a Versioned Data Analysis Environment~\cite{Blomer_poster}}
%
Large-scale scientific endeavors, such as the discovery of the Higgs boson at
the Large Hadron Collider (LHC), often rely on complex software stacks.
Maintaining thousands of dependencies of software libraries and operating system
versions has shown that despite source code availability, the setup and the
validation of a minimal usable analysis environment can easily become
prohibitively expensive. In high-energy physics, a special purpose, open-source,
versioning and snapshotting file system (CernVM-FS) that is used to capture and
distribute entire software stacks, providing instant access to ready-to-run data
analysis environments, proved to be very useful.
% for the global scientific community as well as for
%interested citizens, for instance through the CERN Open Data Portal
%(http://opendata.cern.ch). \todo{revise}

\subsection{Alice Allen: Find it! Cite it!}%~\cite{Allen_poster}}
%
The Astrophysics Source Code Library (ASCL) is an online registry of
scientist-written software used in astronomy research. Their primary interest is
rendering research more transparent by making this software more discoverable for
examination. The ASCL is treated as a publication by an indexing resource for
astronomy, the Astrophysics Data System (ADS). ADS tracks citations to what it
indexes, including citations to ASCL entries. Increasing rewards for writing
software, whether through citation, transitive credit or other methods, gives
software authors a powerful reason to take the time to build sustainability into
their software and is an excellent way to drive community change.

\begin{comment}
The ASCL actively looks for opportunities to use change management strategies to
move astronomy to a more transparent discipline; these strategies can be used in
other disciplines to further change in them.
\end{comment}

>>>>

\begin{enumerate}
\item \textbf{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable
Software through Sustainable Software Architectures~\cite{Venters_poster}.}
Venters proposed that sustainable software is a composite, first-class,
non-functional requirement (NFR) that is at a minimum a measure of a system's
maintainability and extensibility, but may also include other NFRs such as
efficiency (e.g., energy, cost), interoperability, portability, reusability,
scalability, and usability. To achieve technically sustainable software, Venters
suggested that software architectures are fundamental as they are the primary
carrier of system NFRs, i.e., pre-system understanding; and influence how
developers are able to understand, analyze, extend, test, and maintain a
software system, i.e., post-deployment system understanding. In addition,
Venters highlighted that sustainability of software architectures needs to be
addressed to endure different types of change and evolution in order to mitigate
architectural drift, erosion, and knowledge vaporization.

@@

\item \textbf{Marlon Pierce: Patching It Up, Pulling It
Forward~\cite{Pierce_poster}.} Pierce discussed how open open source is. Open
software needs a diverse, openly governed community behind it, just as it needs
open licensing and a public code repository. To probe the level of governance
within open source projects, he and his co-authors (S.~Marru and C.~Mattmann) 
suggested a contest to encourage individual developers to submit
patches and requests to projects that are important to them. This simple
mechanism shall expose several governance mechanisms, such as how easy it is for
independent developers to communicate with project leadership, how projects
accept and license third-party contributions, and how projects make
decisions such as granting source tree write access.

\item \textbf{John Peterson: Continuous Integration for Concurrent MOOSE Framework
and Application Development on GitHub~\cite{Peterson_poster}.}   
Peterson from the Idaho National Laboratory reported
that in March 2014, the MOOSE framework was released under an open source
license on GitHub, significantly expanding and diversifying the pool of current
active and potential future contributors on the project. The MOOSE team employs
an extensive continuous integration test suite to ensure that both the framework
and the applications based on the framework are verified before any code changes
are accepted into the repository. They use a combination of built-in Git
features such as branching and submodules, GitHub API integration capabilities,
and in-house developed testing software to perform this verification and update
the dependent applications in a relatively seamless manner for users.

\item \textbf{Abani Patra: Execute it~\cite{Patra_poster}.}
%Univ at Buffalo, SUNY
Patra discussed the value of an easily accessible platform for
executing scientific software, e.g., HUBzero to access XSEDE or other computing
resources. Such a platform for executing benchmark problems (even at a small
scale) allows the developer community access a reference implementation and
provides an easy way to train the larger user community. A second idea of this
talk was that for true usability, much more attention and support needs
to be given to the integrated use of simulation tools inside complex workflows.

@@

\item \textbf{Daniel Katz: Implementing Transitive Credit with
JSON-LD~\cite{Katz_transitive_credit_poster}.} Science and engineering research
increasingly relies on activities that facilitate research but are not rewarded
or recognized, such as: data sharing; developing common data resources, software
and methodologies; and annotating data and publications. To promote and advance
these activities, we must develop mechanisms for assigning credit, facilitate
the appropriate attribution of research outcomes, devise incentives for
activities that facilitate research, and allocate funds to maximize return on
investment. Katz discussed the issue of assigning credit for both direct and
indirect contributions by using JSON-LD to implement a prototype transitive
credit system.

\item \textbf{Samin Ishtiaq: Daemons, Notifications and Sustaining Software. }%~\cite{Ishtiaq_poster}}
%  Microsoft Research Cambridge
The reproduction and replication of novel results has become a major issue in
computer science, systems biology, and other computational disciplines. These
include both the inability to re-implement novel algorithms and approaches, and
lack of an agreement on how and what to benchmark these algorithms on. Ishtiaq
from Microsoft Research Cambridge pointed out these problems and made several
suggestions to address them.

\item \textbf{James Howison: Retract all Bit-Rotten Publications. }%~\cite{Howison_poster}
Howison sought to provoke discussion by proposing that papers whose workflows
are not kept current with the changing software ecosystem should be
automatically retracted. This would create an incentive for authors to keep
their software current and usable, rather than the current situation in which
every potential user has to do this individually. A softer version of the
proposal would identify papers whose software workflow has become bit-rotten and
allow others to keep the code up to date, either adding them as new authors of
the paper or providing credit in some other form for their academic service.

@@

\item \textbf{Robert Downs: Community Recommendations for Improving Sustainable
Scientific Software Practices~\cite{Downs_poster}.} Robert Downs, of the Columbia
University Center for International Earth Science Information Network (CIESIN),
described a focus group study conducted with the Science Software Cluster (SSC)
of the Federation of Earth Science Information Partners (ESIP). For the study,
almost 300 attendees of the 2014 Summer ESIP Meeting were invited to participate
in simultaneous roundtable discussions on sustainability of science software.
Over two-thirds of the roundtable focus groups responded to a semi-structured
survey that contained three sets of questions eliciting recommendations for
near-term actions of the community to improve sustainable software practices.
Initial analysis of the participants' responses to the questionnaire revealed
several suggestions, which included improving community engagement and
collaborative activities, increasing understanding and awareness, and creating
incentives to motivate sustainable science software practices. The ESIP SSC
plans to engage the community in the recommended activities for improving
sustainable scientific software practices.

\item \textbf{Carl Boettiger: rOpenSci: Building Sustainable Software by Fostering a
Diverse Community~\cite{Boettiger_poster}.}
Boettiger described how the rOpenSci project has been successful by focusing not
just on building software but also on building a community of researchers who
learn and adopt their approaches to reproducible research and sustainable
software practice. Through outreach, mentoring, workshops, and hackathons, they
have not only reached new users, but also turned users into co-developers of
robust software and good practices to support data science research across a
growing set of disciplines.

\item \textbf{Jakob Blomer: The Need for a Versioned Data Analysis Environment~\cite{Blomer_poster}.}
Large-scale scientific endeavors, such as the discovery of the Higgs boson at
the Large Hadron Collider (LHC), often rely on complex software stacks.
Maintaining thousands of dependencies of software libraries and operating system
versions has shown that despite source code availability, the setup and the
validation of a minimal usable analysis environment can easily become
prohibitively expensive. In high-energy physics, CernVM-FS, a special-purpose, open-source,
versioning, and snapshotting file system used to capture and
distribute entire software stacks, proved to be useful for providing instant access to ready-to-run data
analysis environments.
% for the global scientific community as well as for
%interested citizens, for instance through the CERN Open Data Portal
%(http://opendata.cern.ch). \todo{revise}

\item \textbf{Alice Allen: Find it! Cite it! }%~\cite{Allen_poster}}
The Astrophysics Source Code Library (ASCL) is an online registry of
scientist-written software used in astronomy research. Their primary interest is
rendering research more transparent by making this software more discoverable for
examination. The ASCL is treated as a publication by an indexing resource for
astronomy, the Astrophysics Data System (ADS). ADS tracks citations to what it
indexes, including citations to ASCL entries. Increasing rewards for writing
software, whether through citation, transitive credit or other methods, gives
software authors a powerful reason to take the time to build sustainability into
their software and is an excellent way to drive community change.
\begin{comment}
The ASCL actively looks for opportunities to use change management strategies to
move astronomy to a more transparent discipline; these strategies can be used in
other disciplines to further change in them.
\end{comment}
\end{enumerate}

\smallskip
